Quantifying Uncertainty in Forecast-Based Anticipatory Action Triggers Using Bootstrapping Methods

### 2025-08-22 paper.tex
Some small text tweaks and updating of comments

\subsection{The PyCPT seasonal forecasting method}

% Max to write but could probably use Kevin and Nathan's eyes to make sure I'm not saying anything scientifically out of pocket.
% Dan style note: "PyCPT" instead of "NextGen" 
% Talk about heuristic rules that the country offices were using, and how they are similar conceptually / functionallty to bootstrapping

Talk about the types of seasonal forecasts we are using here, where they come from and how to interpret them. Can draw from a lot of prior IRI work here. Key talking points are that NextGen is a method to produce downscaled and bias-corrected seasonal forecasts using a multi-model ensemble of GCMs. The models, predictand and calibration parameters can be customized to each locality and context. Local met services use the method and software to create purpose-built forecast models for their country or region. Forecasts are presented as the probability of non-exceedance at a given percentile threshold (not only terciles).

@@

\section{Sources of forecast uncertainty}

% Nathan - can fold most of this into section 2b, since it isn't really a standalone section

% Kevin to write w Max and Nathan as needed.
% The basic strategy is to go through all of the different commonly discussed sources of uncertainty, and talk about how they are addressed in the NextGen approach, except for the problem of the limited historical record. That is what we're addressing with this paper.

% note Aug 13: Can pare down the discussion of the first four sources of uncertainty, since they aren't the main point.

\subsection{Uncertainty in modeling choices}

Different GCMs have different assumptions about how things work. We account for this by using a multi-model ensemble in predictions.

\subsection{Uncertainty in initial conditions}

I'm not sure if or how to talk about this -- I'm mostly familiar with this in the context of long-term climate change projections, not seasonal forecasting. As I understand it, all of the GCM output that we use for seasonal forecasting is initialized with the same conditions for a given issue month. So maybe this is not a huge issue the way it is for CC projections, where uncertainty in initial conditions can propagate further (cf Schwarzwald and Lenssen 2022). But maybe I'm also missing something, so it would be good to get your input.

In any case, we can talk about how we use forecasts at different lead times, with the decision rule putting greater weight on forecasts with a shorter lead time, since we are more certain about the initial conditions.

\subsection{Uncertainty in accuracy of predictions}

The bias correction and calibration components of NextGen are meant to account for this. I need to bug Andy R. and Azhar about this since they have promised a more thorough writeup. Talk about the sharpness ratio of quantifying the confidence of the forecast (this is something that decision-makers should also be thinking about, and not only whether the forecast is above or below the threshold).

Note that this section is mostly focusing on how well forecasts predict observed rainfall. There is a second-order question about whether forecasted rainfall is related to humanitarian impacts of drought (e.g. hunger, loss of animals), which is of practical importance to the projects, but not really what this paper is about. We may want to at least mention it in passing.

@@

% Nathan to write w Kevin and Max and needed

% MAX TODO: Make sure we are using block randomization, with consecutive years. This could help deal with the autocorrelation problem. There is a literature we can tie this to

% Abdel is particularly interested in the question of whether interannual correlation and / or trends make the calculation of confidence intervals using this method problematic. i need to put on my stats hat this this one.

% Note that if we were really doing this properly, I think we would want to do the bootstrap sampling during the step of per-pixel model calibration (which happens in CPT), rather than doing it after the fact with areal average data as we are doing here. But that is a matter for a future paper. We want to at least mention it though I think.

% kevin note: we are thinking about this as applicable to things beyond PyCPT, tho, so dont get too hung up about this

@@

To test how sensitive the protocol is to the exact threshold values chosen, this section presents an uncertainty analysis. This analysis is based on the method of re-estimating the trigger threshold with years randomly omitted from the historical record. Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error.

@@

% Need to be sure we are using a definition of the percentile that is consistent with what the API uses! R's may not be the same. make sure we are consistent

@@

% tweak visual so that user focuses on the boxplot, not the clim odds

% Note: The figures seem to get added to the end instead of in-line. I'm guessing that's how it is in the journal?

% note that this method can acommodate compound / multimonth rules

>>>>

\subsection{The CPT seasonal forecasting method}

% Max to write but could probably use Kevin and Nathan's eyes to make sure I'm not saying anything scientifically out of pocket.
% Talk about heuristic rules that the country offices were using, and how they are similar conceptually / functionallty to bootstrapping

Talk about the types of seasonal forecasts we are using here, where they come from and how to interpret them. Can draw from a lot of prior IRI work here. Key talking points are that CPT is a method to produce downscaled and bias-corrected seasonal forecasts using a multi-model ensemble of GCMs. The models, predictand and calibration parameters can be customized to each locality and context. Local met services use the method and software to create purpose-built forecast models for their country or region. Forecasts are presented as the probability of non-exceedance at a given percentile threshold (not only terciles).

Talk a little bit about how the CPT method accounts for observational and model uncertainty.

@@

@@

% Nathan to write w Kevin and Max and needed. Tag in Abdel to see if this addresses his concerns about correlation

@@

To test how sensitive the protocol is to the exact threshold values chosen, this section presents an uncertainty analysis. This analysis is based on the method of re-estimating the trigger threshold with years randomly omitted from the historical record. Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error. MAKE SURE WE TALK ABOUT BLOCK RANDOMIZATION HERE

@@

@@

% note that this method can acommodate compound / multimonth rules - be sure to mention that
